Supplementary Materials:\\Evidence for black-swan events in animal populations

### 2016-01-02 text/ms.tex
edits

\textbf{Abstract}:
Black swans are statistically improbable events that nonetheless occur---often with profound consequences\cite{taleb2007, sornette2009}. While extremes in the physical environment, such as monsoons and heat waves, are widely studied and increasing in magnitude and frequency\cite{meehl2004, katz2005, ipcc2012}, it remains unclear the extent to which ecological populations buffer or suffer from such extremes. Here we derive a model for estimating the frequency and direction of black-swan events and apply it to abundance time series of 609 animal populations. We find strong evidence for black-swan events, but they are rare, occurring in \(\sim\)\overallBasePerc \% of populations: most frequently for birds (\birdPH \%), mammals (\mammalsPH \%), and insects (\insectsPH \%). Black-swan events are predominantly (\percBSDown \%) downward, implying that unexpected population crashes occur far more frequently than unexpected population increases. Black-swan events are not explained by any intrinsic life-history covariates, but tend to be driven by extrinsic perturbations such as climate or severe winters. Data and simulations reveal that black-swan events are more likely to be detected in longer and less noisy time series. Rare events tend to be crashes and ignoring them will underestimate extinction risk. Ignoring black-swan events leads to a \crashUnderMedian-fold (50\% interval \crashUnderRange) underestimate in the magnitude of population crashes. Our results demonstrate the critical importance of allowing for downward and heavy-tailed events in population modelling and developing robust conservation and management strategies, particularly in the face of increased climate extremes.

\bigskip
\bigskip

\textbf{One Sentence Summary}:

@@

\textbf{Acknowledgements}: We thank J.W. Moore, A.O. Mooers, L.R. Gerber, J.D. Yeakel, C. Minto and members of the Earth to Ocean Group for helpful discussions and comments. We are grateful to the contributors and maintainers of the Global Population Dynamics Database and to Compute Canada's WestGrid high-performance computing resources. Silhouette images were obtained from phylopic.org under Creative Commons licenses; sources are listed in the Supplementary Information. Funding was provided by a Simon Fraser University Graduate Fellowship (S.C.A.), the Natural Sciences and Engineering Research Council of Canada (N.K.D., A.B.C.), the Canada Research Chairs Program (N.K.D.).

>>>>

\textbf{Abstract}:
Black swans are improbable events that nonetheless occur---often with profound
consequences. While physical extremes, such as monsoons and heat waves, are
widely studied and increasing in magnitude and frequency, it remains unclear
the extent to which ecological populations buffer or suffer from such
extremes. Here we estimate the frequency and direction of black-swan events in
609 animal populations. We find strong evidence for black-swan events
occurring in \(\sim\)\overallBasePerc \% of populations: most frequently for
birds (\birdPH \%), mammals (\mammalsPH \%), and insects (\insectsPH \%).
Black-swan events are predominantly population crashes (\percBSDown \%),
driven by extrinsic perturbations such as climate or severe winters, and
ignoring them may underestimate extinction risk. Our results demonstrate the
importance of allowing for heavy-tailed downward events in population
modelling and developing robust conservation strategies, particularly in the
face of increased climate extremes.

\textbf{One Sentence Summary}: Black swans---statistically improbable events
with profound consequences---can and do occur in animal populations and
ignoring them may severely underestimate extinction risk.

%\clearpage

@@

\textbf{Acknowledgements}: We thank J.W. Moore, A.O. Mooers, L.R. Gerber, J.D. Yeakel, C. Minto and members of the Earth to Ocean Group for helpful discussions and comments. We are grateful to the contributors and maintainers of the Global Population Dynamics Database and to Compute Canada's WestGrid high-performance computing resources. Silhouette images were obtained from phylopic.org under Creative Commons licenses; sources are listed in the Supplementary Materials. Funding was provided by a Simon Fraser University Graduate Fellowship (S.C.A.), the Natural Sciences and Engineering Research Council of Canada (N.K.D., A.B.C.), the Canada Research Chairs Program (N.K.D.).

### 2016-01-02 text/supp.tex
edits

\title{Supplementary Information:\\Evidence for black-swan events in animal
  populations}

@@

We selected abundance time series from the Global Population Dynamics Database\cite{gpdd2010} (GPDD), which contains nearly 5,000 time series of abundance from \(\sim\) 1,000 species and \(\sim\) 100 taxonomic orders. We filtered the data (Supplementary Information) to remove populations from less reliable data sources, and those without sufficient data for our models, and then interpolated some missing values (sensu Brook \textit{et al.}\cite{brook2006a}). Our interpolation affected only \(\sim\) \interpPointsPerc \% of the final data points (table~S1) and none of the data points that were later considered black-swan events. Our final dataset contained \NPops\ populations across \NOrders\ taxonomic orders and seven taxonomic classes, with a median of \medianTimeSteps\ time steps (range \minTimeSteps--\maxTimeSteps) (table~S1).

@@

There are a number of caveats when considering the generality of our results. The GPDD data represent a taxonomically and geographically biased sample of populations---the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom. Although we would expect to find qualitatively similar evidence for black swans in other large taxonomic or geographic samples of populations, the common forces driving those black swans would likely differ. Additionally, some apparent black-swan events could be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes. Indeed, we discarded three of the populations that our method initially identified as heavy tailed because they turned out to be data-entry errors (Supplementary Information). Finally, the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and these likely influence the detection of heavy tails. For example, if we make frequent observations relative to generation time (e.g.~for many large-bodied mammals) we will average across generations and perhaps miss black-swan events. Conversely, if we census populations infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and also dampen black-swan dynamics.

@@

\section{Simulation testing} We performed two types of simulation testing (Supporting Information). First, we tested how easily the Student-t distribution \(\nu\) parameter could be recovered given different true values of \(\nu\) and different sample sizes. Second, we tested the ability of the heavy-tailed Gompertz model to obtain unbiased parameter estimates of \(\nu\) given that a set of process deviations was provided in which the effective \(\nu\) value was close to the true \(\nu\) value.

@@

\section{Skewed Student-t forecasts} To formally evaluate the apparent skewness of heavy-tailed process noise we fit Gompertz models with skewed Student-t distributed process noise\cite{fernandez1998} (Supplementary Information). The distribution adds one parameter to the Student-t distribution, \(\gamma\), which controls the skewness. The distribution is symmetrical if \(\gamma = 1\), left skewed if \(0 < \gamma < 1\), and right skewed if \(1 < \gamma < \infty\). We placed a weakly informative prior of \(\mathrm{Cauchy}(0, 2.5)\) on \(\log \gamma\). We aggregated 5,000 randomly selected posterior samples from the \(\gamma\) parameter of each model at three levels of evidence for heavy-tails: \emph{heavy tailed} median \(\nu < 10\); \emph{slightly heavy tailed} \(10 \le \mathrm{median}\, \nu \le 70\), and \emph{normal tails}, median \(\nu > 70\) (Fig.~4).

>>>>

\title{Supplementary Materials:\\Evidence for black-swan events in animal
  populations}

@@

We selected abundance time series from the Global Population Dynamics Database\cite{gpdd2010} (GPDD), which contains nearly 5,000 time series of abundance from \(\sim\) 1,000 species and \(\sim\) 100 taxonomic orders. We filtered the data to remove populations from less reliable data sources, and those without sufficient data for our models, and then interpolated some missing values (sensu Brook \textit{et al.}\cite{brook2006a}). Our interpolation affected only \(\sim\) \interpPointsPerc \% of the final data points (table~S1) and none of the data points that were later considered black-swan events. Our final dataset contained \NPops\ populations across \NOrders\ taxonomic orders and seven taxonomic classes, with a median of \medianTimeSteps\ time steps (range \minTimeSteps--\maxTimeSteps) (table~S1).

@@

There are a number of caveats when considering the generality of our results. The GPDD data represent a taxonomically and geographically biased sample of populations---the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom. Although we would expect to find qualitatively similar evidence for black swans in other large taxonomic or geographic samples of populations, the common forces driving those black swans would likely differ. Additionally, some apparent black-swan events could be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes. Indeed, we discarded three of the populations that our method initially identified as heavy tailed because they turned out to be data-entry errors. Finally, the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and these likely influence the detection of heavy tails. For example, if we make frequent observations relative to generation time (e.g.~for many large-bodied mammals) we will average across generations and perhaps miss black-swan events. Conversely, if we census populations infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and also dampen black-swan dynamics.

@@

\section{Simulation testing} We performed two types of simulation testing. First, we tested how easily the Student-t distribution \(\nu\) parameter could be recovered given different true values of \(\nu\) and different sample sizes. Second, we tested the ability of the heavy-tailed Gompertz model to obtain unbiased parameter estimates of \(\nu\) given that a set of process deviations was provided in which the effective \(\nu\) value was close to the true \(\nu\) value.

@@

\section{Skewed Student-t forecasts} To formally evaluate the apparent skewness of heavy-tailed process noise we fit Gompertz models with skewed Student-t distributed process noise\cite{fernandez1998}. The distribution adds one parameter to the Student-t distribution, \(\gamma\), which controls the skewness. The distribution is symmetrical if \(\gamma = 1\), left skewed if \(0 < \gamma < 1\), and right skewed if \(1 < \gamma < \infty\). We placed a weakly informative prior of \(\mathrm{Cauchy}(0, 2.5)\) on \(\log \gamma\). We aggregated 5,000 randomly selected posterior samples from the \(\gamma\) parameter of each model at three levels of evidence for heavy-tails: \emph{heavy tailed} median \(\nu < 10\); \emph{slightly heavy tailed} \(10 \le \mathrm{median}\, \nu \le 70\), and \emph{normal tails}, median \(\nu > 70\) (Fig.~4).
